---
title: "Multivariate Contrasts"
author: "David Lieberman, Yavuz Ramiz Çolak, Ryo Tamaki, Liana Wang"
date: "April 5th, 2019"
output:
  html_document: default
---

Environmental Attitude Survey:

For four questions, the scale is (1=Strongly Agree, 5=Strongly Disagree)
v10 : Modern science will solve our environmental problems with little change to our way of life
v12 : Almost everything we do in modern life harms the environment.
V25 : There is no point in doing what I can for the environment unless others do the same
v47 : Poorer countries should be expected to make less effort than richer countries to protect the environment

One question has a different scale :
v38 : In general, do you think that a rise in the world's temperature caused by the greenhouse effect is (1=Extremely Dangerous for you and family, 5=Not dangerous for you and family)

In addition, the following variables were measured
V3 : Country (Japan, US, Mexico, Netherlands)
V200 : Gender (1=male, 2=female)
V201 : Age in Years
V202 : Marital Status (1=Married, 2=Widowed, 3=Divorced, 4=Separated, 5=Single never married)
V204 : Years of education
V205 : Highest Degree obtained
V246 : Political Views (1=Extremely Liberal (Communist), 5=Extremely Conservative (Fascist))

```{r}
envsurvey <-  read.csv("http://www.reuningscherer.net/stat660/data/Environmental_Survey_MANOVA.csv", header=T)
envsurvey <- envsurvey[1:6]
envsurvey
```

```{r}
multicontrast <- function(contrast, data, grouping){
  
  # Groups
  groups <- as.vector(as.matrix(unique(grouping)))
  
  # Multivariate Means for Each Variable in Each Group
  M <- apply(data, 2, function(y) tapply(y, grouping, mean))
  M <- M[match(groups, row.names(M)),]
  
  # Counts for Each Group
  N <- table(grouping)
  N <- N[match(groups, row.names(N))]
  
  # Calculate Weighted Sum of Squared Weights
  SW <- sum(contrast^2 / N)
  
  # Calculate SSCP Between (Hypothesis Matrix)
  C_hat <- colSums(contrast*M)
  SSCP_between <- (cbind(C_hat) %*% rbind(C_hat)) / SW
  
  # Calculate SSCP Within (Error Matrix)
  SSCP_within_each_group <- list()
  for (i in seq_along(groups)){
    X <- subset(data, grouping == groups[i])
    deviations <- matrix(0, nrow(X), ncol(X))
    for (j in 1:ncol(X)){
      deviations[,j] <- X[,j] - M[i,j]
    }
    SSCP_within_each_group[[i]] <- t(deviations) %*% deviations
  }
  SSCP_within <- Reduce("+", SSCP_within_each_group)
  
  # Calculate Wilks' Lambda
  lambda <- det(SSCP_within) / det(SSCP_between + SSCP_within)
  
  # Calculate Degrees of Freedom
  p <- ncol(data) # no.variables
  m <- nrow(data) - length(groups)  # no.observations - no.groups
  
  df1 <- p
  df2 <- m-p+1
  
  # Calculate approx. F
  F.stat <- (1 - lambda) / lambda * df2 / df1
  
  # Calculate p-value from F distribution
  p.value <- 1 - pf(F.stat, df1, df2)
  
  # Output
  out <- c(lambda, F.stat, df1, df2, p.value)
  names(out) <- c("Wilks", "approx.F", "df1", "df2", "p.value")
  return(out)
}
```

```{r}
as.matrix(unique(envsurvey[1])) # Just to see the countries' order for the contrasts vector
```

Japan vs US
```{r}
multicontrast(c(1,0,-1,0), envsurvey[2:6], envsurvey[1])
```

US vs Rest of the World
```{r}
multicontrast(c(3,-1,-1,-1), envsurvey[2:6], envsurvey[1])
```